---
title: "Homework 1"
author: "Benjamin Zeiger"
format: 
  html:
    code-overflow: wrap
    fig-width: 8
    fig-height: 6
---

## Question 1

Determine how many different genotypes are present at a locus with n alleles that differ by state. You already know that there is one genotype at a locus with one allele and three genotypes at a locus with two alleles. Continue this with three, four, and more alleles until you derive the general case. Then create a graph showing how the number of genotypes varies by the number of alleles at a diploid locus.

## Answer

In a one-allele, diploid gene, there is only one option: AA In a two-allele, diploid gene, there will be three options: AA, plus two new ones: BB and AB. In a three-allele gene, there are now six genotypes: all the previous options, plus three new ones: AC, BC, and CC

The number of possible genotypes increase following the triangle numbers (1, 3, 6, 10), so the number of possible genotypes in a system of *n* alleles is equal to the sum of all integers between 1 and *n*. This sum is also equal to *n(n+1)/2*, which makes it easier to generate a vector, but I forgot how to derive that from the summation, so I did have to reference the wikipedia on "Triangular Numbers" to get here.

Graphed below, we see the number of possible genotypes increases exponentially as the number of alleles increases

```{r}

library(tidyverse)

alleles = seq(1,10,1)
genotypes = alleles*(alleles+1)/2

ggplot(mapping = aes(x = alleles, y = genotypes)) +
  geom_point()

```

## Question 2

Graph the frequencies of the heterozygotes as a function of the allele frequencies ‘p’ & ‘q’ in a three ALLELE Hardy Weinberg Equilibrium system. At what allele frequencies are the frequency of heterozygotes maximized?

## Answer

If we allow 'p' to represent the frequency of allele 'A,' and 'q' to represent frequency of allele 'B,' then our 3^rd^ value will be 'r' and will represent the frequency of allele 'C.'

In order to represent this in terms of 'p' and 'q' we will let r = 1 - p - q.

The frequencies of each genotype should then be

f~AA~ = p^2^

f~BB~ = q^2^

f~CC~ = (1-p-q)^2^

f~AB~ = pq

f~AC~ = p \*

```{r}



```

## Question 3

In a sample of 1,617 Spanish Basques, the numbers of A, B, O, and AB blood types observed were 724, 110, 763, and 20, respectively. These blood groups are due to three alleles, IA, IB, and IO, with IAIAand IAIO having blood group A; IBIB and IBIO having blood group B, IAIB having blood group AB, and IOIOhaving blood group O. The best estimates of the alleles frequencies in Basque sample are 0.2661 for IA, 0.0411 for IB, and 0.6928 for IO. Calculate the expected numbers of the four blood group phenotypes and carry out a chi-square test for HWE.

```{r}



```
