---
title: "EEOB 5330 -- Spring 2026"
author: "Lecture 11 - February 18, 2026"
date: 'Topic: Selection'
output:
  beamer_presentation:
    fig_width: 1.5
    fig_height: 1.5
    fig_caption: no
    keep_tex: yes
    includes:
      in_header: preamble.tex
  ioslides_presentation: default
fontsize: 10pt
---

```{r, echo=FALSE, tidy=TRUE, eval=TRUE, mysize=TRUE, size='\\scriptsize', fig.align="center", fig.width=3.8, fig.height=2.8, comment=""}
library("knitr") ## if this gives an error, run first install.packages("knitr")
options(stringsAsFactors=FALSE)
knitr::knit_hooks$set(mysize = function(before, options, envir) {
  if (before) 
    return(options$size)
})
options(width = 60)
set.seed(1)
```

## \small Review and plan for today

\begin{itemize}
\item {\ncr Last time:} Introduction to modeling selection
\myvs
\item What assumptions do we make in HWE?
\begin{itemize}
\item  {\ncb no mutation} - no new alleles are introduced via point mutations
\item {\bf {\ncb no selection} - alleles are not adaptive}
\item  {\ncb no migration} - no new alleles are introduced via the movement of individuals
\item   {\ncb no genetic drift} - population size is infinitely large (or nearly so)
\item  {\ncb no assortative mating} - each individual has an equal chance of producing
offspring with any member of the opposite sex
\end{itemize}
\myvs
\item {\ncr Today:} Look at types of selection and their effect on alleles frequencies; Tajima's D
\end{itemize}
\vspace{0.5in}

## \small Modeling selection

\begin{itemize}
\item {\ncr Idea:} Ignore mutation and drift, and think about how to add selection to the formation of a new generation
\item Consider a single-locus, 2 allele system, with $p$ = frequency of the $A$ allele, $q=1-p$ = frequency of the $a$ allele
\end{itemize}
\myvs
\begin{center}
\begin{tabular}{c c c c}
\hline
& & & \\
{\ncb Genotype} & {\bf $AA$} & {\bf $Aa$} & {\bf $aa$}\\
& & & \\
\hline
& & & \\
{\ncb Frequency} & $p^2$ & $2pq$ & $q^2$\\
& & & \\
{\ncb Viability} & $w_{AA}$ & $w_{Aa}$ & $w_{aa}$ \\
& & & \\
{\ncb Frequency} & $\frac{p^2 w_{AA}}{\bar{w}}$ & $\frac{2pqw_{Aa}}{\bar{w}}$ & $\frac{q^2 w_{aa}}{\bar{w}}$ \\
{\ncb after selection} & & & \\
& & & \\
\hline
\end{tabular}
\end{center}
\myvs\myvs

where $\bar{w} = p^2 w_{AA} + 2pq w_{Aa} + q^2 w_{aa}$ \vspace{0.5in}

## \small Modeling selection

\begin{itemize}
\item {\ncr Question:} How does the frequency of the $A$ allele change in one generation?
\vspace{0.5in}
\item After some math, we have
$$ p^{'} - p = \Delta p = \frac{pq \left[ p(w_{AA} - w_{Aa}) + q(w_{Aa}-w_{aa})\right]}{p^2 w_{AA} + 2pq w_{Aa} + q^2 w_{aa}}$$
\item {\ncb Gillespie:} ``{\itshape single most important equation in all of population genetics and evolution}''
\myvs
\item {\ncb Why?} Allows us to study dynamics of a system under selection by looking at relationships between genotype viabilities
\end{itemize}
\vspace{0.5in}

## \small Relative vs. absolute fitness

\begin{itemize}
\item Note that the model doesn't change if we instead view \begin{color}{green}viabilities\end{color} as being {\ncr relative} to one another (which is all we can typically do in practice)
\end{itemize}
\myvs\myvs
\begin{center}
\begin{tabular}{ c c c c}
 & $\mathbf{AA}$ & $\mathbf{Aa}$ & $\mathbf{aa}$\\
 & & & \\
\begin{color}{green}viability\end{color} & \begin{color}{green}$w_{AA}$\end{color} & \begin{color}{green}$w_{Aa}$\end{color} & \begin{color}{green}$w_{aa}$\end{color} \\
& & & \\
{\ncr relative fitness} & {\ncr 1} & {\ncr $w_{Aa}/w_{AA}$} & {\ncr $w_{aa}/w_{AA}$} \\
& & & \\
{\ncb More intuitive} & {\ncb 1} & {\ncb $1-hs$} & {\ncb $1-s$} \\
{\ncb parameterization} & & & \\

\end{tabular}
\end{center}
\myvs
\begin{itemize}
\item  $s$ = selection coefficient:
\myvs
\item $h$ = heterozygous effect:
\end{itemize}
\vspace{0.25in}

## \small Heterozygous effect

\begin{itemize}
\item $h$ = measure of the fitness of the heterozygote relative to the selective difference between the two homozygotes
\myvs
\item Interpretation:
\myvs
\begin{center}
\begin{tabular}{c c}
$h=0$ & $A$ dominant, $a$ recessive\\
$h = 1$ & $a$ dominant, $A$ recessive\\
$0<h<1$ & incomplete dominance \\
$h<0$ & overdominance \\
$h>1$ & underdominance\\
\end{tabular}
\end{center}
\myvs
\item We will look at the dynamics of the allele frequency $p$ for each of these cases
\myvs
\item Note that $\Delta p$ becomes
$$\Delta p = \frac{pqs \left[ ph+q(1-h)\right]}{\bar{w}}$$
where $$\bar{w} = 1-2pqhs-q^2 s$$
\end{itemize}
\vspace{0.25in}

## \small Case 1: directional selection

\begin{itemize}
\item {\ncr Directional selection} refers to the case of {\ncb incomplete dominance:}  $0<h<1$
\myvs
\begin{itemize}
\item fitness of $AA$ > fitness of $Aa$ > fitness of $aa$
\end{itemize}
\myvs
\item {\ncr Question:} How do we expect the frequency of the $A$ allele to change over time?
\vspace{2.0in}
\end{itemize}

## \small Case 1: directional selection

\begin{itemize}
\item Graph $\Delta p$ as a function of $p$ for $h=0.5$ and $s=0.1$
\end{itemize}

```{r, echo=F, eval=T, fig.align="center", fig.width=4.25, fig.height=3.5}
myseq=seq(0,1,0.01)
deltap = function(pA,h,s){
  wbar =  1-2*pA*(1-pA)*h*s-s*(1-pA)^2
  deltapA = pA*(1-pA)*s*(pA*h+(1-pA)*(1-h))/wbar
  return(deltapA)
}
plot(myseq,deltap(myseq,0.5,0.1),xlab="allele frequency, p",ylab=expression(paste(Delta," p")),lwd=3,type='l', col="maroon",ylim=c(0,0.015))
```

## \small Case 1: directional selection

\begin{itemize}
\item Graph the value of $p$ in the next generation as a function of the current value of $p$ for $h=0.5$ and $s=0.1$
\end{itemize}

```{r, echo=F, eval=T, fig.align="center", fig.width=3.25, fig.height=3.5}
p.by.gen=matrix(0,ncol=2,nrow=101)
p.by.gen[1,1] = 0
p.by.gen[1,2] = 0.05
for (i in 2:101) {
  p.by.gen[i,1] = i-1
  p.by.gen[i,2] = p.by.gen[i-1,2] + deltap(p.by.gen[i-1,2],0.5,0.1)
}
plot(p.by.gen[,1],p.by.gen[,2],xlab="Generation",ylab="p",lwd=3,type='l', col="lightskyblue",ylim=c(0,1))
```

## \small Case 1: directional selection

\begin{itemize}
\item Regardless of the starting point for $p$, the frequency will always move toward 1
\begin{itemize}
\item Will it ever reach 1?
\end{itemize}
\vspace{0.5in}
\item The value $p=1$ is called a {\ncr stable equilibrium}
\myvs
\item The value $p=0$ is called {\ncb unstable equilibrium}
\myvs
\end{itemize}
\vspace{0.5in}

## \small Case 2: balancing selection

\begin{itemize}
\item {\ncr Balancing selection} refers to the case of {\ncb overdominance:}  $h<0$
\myvs
\begin{itemize}
\item fitness of $Aa$ > $AA$ and fitness of $Aa$ > $aa$
\end{itemize}
\myvs
\item {\ncr Question:} How do we expect the frequency of the $A$ allele to change over time?
\vspace{2.0in}
\end{itemize}

## \small Case 2: balancing selection

\begin{itemize}
\item Graph $\Delta p$ as a function of $p$ for $h=-0.5$ and $s=0.1$
\end{itemize}

```{r, echo=F, eval=T, fig.align="center", fig.width=4.25, fig.height=3.5}
myseq=seq(0,1,0.01)
deltap = function(pA,h,s){
  wbar =  1-2*pA*(1-pA)*h*s-s*(1-pA)^2
  deltapA = pA*(1-pA)*s*(pA*h+(1-pA)*(1-h))/wbar
  return(deltapA)
}
plot(myseq,deltap(myseq,-0.5,0.1),xlab="allele frequency, p",ylab=expression(paste(Delta," p")),lwd=3,type='l', col="maroon",ylim=c(-0.01,0.02))
lines(c(-0.05,1.05),c(0,0))
```

## \small Case 2: balancing selection

\begin{itemize}
\item Graph the value of $p$ in the next generation as a function of the current value of $p$ for $h=-0.5$ and $s=0.1$
\end{itemize}

```{r, echo=F, eval=T, fig.align="center", fig.width=3.25, fig.height=3.5}
p.by.gen=matrix(0,ncol=2,nrow=101)
p.by.gen[1,1] = 0
p.by.gen[1,2] = 0.1
p.by.gen2=matrix(0,ncol=2,nrow=101)
p.by.gen2[1,1] = 0
p.by.gen2[1,2] = 0.9
for (i in 2:101) {
  p.by.gen[i,1] = i-1
  p.by.gen[i,2] = p.by.gen[i-1,2] + deltap(p.by.gen[i-1,2],-0.5,0.1)
  p.by.gen2[i,1] = i-1
  p.by.gen2[i,2] = p.by.gen2[i-1,2] + deltap(p.by.gen2[i-1,2],-0.5,0.1)
}
plot(p.by.gen[,1],p.by.gen[,2],xlab="Generation",ylab="p",lwd=3,type='l', col="lightskyblue",ylim=c(0,1.05))
lines(p.by.gen2[,1],p.by.gen2[,2],col="lightskyblue",lwd=3)
```

## \small Case 2: balancing selection

\begin{itemize}
\item {\ncr Stable equilibrium} at $\tilde{p} = \frac{h-1}{2h-1}$
\myvs
\item {\ncb Unstable equilibria} at $p=0$ and $p=1$
\myvs
\item Empirical example:  sickle-cell disease and malaria
\myvs
\begin{itemize}
\item Heterozygotes do not have sickle-cell disease {\itshape and} are resistant to malaria
\item Homozygotes for the non-sickle-cell allele are susceptible to malaria
\item Homozygotes for the sickle-cell allele have sickle-cell disease
\item Interestingly, allele frequencies vary in regions where malaria is less prevalent or absent, reflecting differing selective pressures
\end{itemize}
\end{itemize}
\vspace{0.5in}

## \small Case 3: disruptive selection

\begin{itemize}
\item {\ncr disruptive selection} refers to the case of {\ncb underdominance:}  $h>1$
\myvs
\begin{itemize}
\item fitness of $Aa$ < $AA$ and fitness of $Aa$ < $aa$
\end{itemize}
\myvs
\item {\ncr Question:} How do we expect the frequency of the $A$ allele to change over time?
\vspace{2.0in}
\end{itemize}

## \small Case 3: disruptive selection

\begin{itemize}
\item Graph $\Delta p$ as a function of $p$ for $h=1.5$ and $s=0.1$
\end{itemize}

```{r, echo=F, eval=T, fig.align="center", fig.width=4.25, fig.height=3.5}
myseq=seq(0,1,0.01)
deltap = function(pA,h,s){
  wbar =  1-2*pA*(1-pA)*h*s-s*(1-pA)^2
  deltapA = pA*(1-pA)*s*(pA*h+(1-pA)*(1-h))/wbar
  return(deltapA)
}
plot(myseq,deltap(myseq,1.5,0.1),xlab="allele frequency, p",ylab=expression(paste(Delta," p")),lwd=3,type='l', col="maroon",ylim=c(-0.01,0.02))
lines(c(-0.05,1.05),c(0,0))
```

## \small Case 3: disruptive selection

\begin{itemize}
\item Graph the value of $p$ in the next generation as a function of the current value of $p$ for $h=1.5$ and $s=0.1$
\end{itemize}

```{r, echo=F, eval=T, fig.align="center", fig.width=3.25, fig.height=3.5}
p.by.gen=matrix(0,ncol=2,nrow=101)
p.by.gen[1,1] = 0
p.by.gen[1,2] = 0.1
p.by.gen2=matrix(0,ncol=2,nrow=101)
p.by.gen2[1,1] = 0
p.by.gen2[1,2] = 0.9
p.by.gen3=matrix(0,ncol=2,nrow=101)
p.by.gen3[1,1] = 0
p.by.gen3[1,2] = 0.25
p.by.gen4=matrix(0,ncol=2,nrow=101)
p.by.gen4[1,1] = 0
p.by.gen4[1,2] = 0.26
p.by.gen5=matrix(0,ncol=2,nrow=101)
p.by.gen5[1,1] = 0
p.by.gen5[1,2] = 0.24
for (i in 2:101) {
  p.by.gen[i,1] = i-1
  p.by.gen[i,2] = p.by.gen[i-1,2] + deltap(p.by.gen[i-1,2],1.5,0.1)
  p.by.gen2[i,1] = i-1
  p.by.gen2[i,2] = p.by.gen2[i-1,2] + deltap(p.by.gen2[i-1,2],1.5,0.1)
  p.by.gen3[i,1] = i-1
  p.by.gen3[i,2] = p.by.gen3[i-1,2] + deltap(p.by.gen3[i-1,2],1.5,0.1)
  p.by.gen4[i,1] = i-1
  p.by.gen4[i,2] = p.by.gen4[i-1,2] + deltap(p.by.gen4[i-1,2],1.5,0.1)
  p.by.gen5[i,1] = i-1
  p.by.gen5[i,2] = p.by.gen5[i-1,2] + deltap(p.by.gen5[i-1,2],1.5,0.1)
}
plot(p.by.gen[,1],p.by.gen[,2],xlab="Generation",ylab="p",lwd=3,type='l', col="lightskyblue",ylim=c(0,1.05))
lines(p.by.gen2[,1],p.by.gen2[,2],col="lightskyblue",lwd=3)
lines(p.by.gen3[,1],p.by.gen3[,2],col="lightskyblue",lwd=3)
lines(p.by.gen4[,1],p.by.gen4[,2],col="lightskyblue",lwd=3)
lines(p.by.gen5[,1],p.by.gen5[,2],col="lightskyblue",lwd=3)
```

## \small Case 3: disruptive selection

\begin{itemize}
\item {\ncr Stable equilibria} at $p=0$ and $p=1$ (but only locally -- not globally)
\myvs
\item {\ncb Unstable equilibrium} at $\tilde{p} = \frac{h-1}{2h-1}$
\myvs
\item Empirical examples are rare
\end{itemize}
\vspace{1.0in}

## \small Mean fitness

\begin{itemize}
\item Sewall Wright dervied the following to unify these possibilities:
\myvs
{\ncr $$\Delta p = \frac{pq}{2\bar{w}}\frac{d\bar{w}}{dp}$$}
\item Some consequences:
\myvs
\begin{itemize}
\item the allele frequency always changes so that mean fitness increases

\item the rate of change in $p$ is proportional to the genetic variation in the population ($pq$)\\
\end{itemize}
\myvs\myvs
 {\ncb $\implies$ selection always increases the mean fitness of the population, and does so at a rate proportional to the genetic variation}
\end{itemize}
\vspace{0.5in}

## \small Mutation-selection balance

\begin{itemize}
\item Expect a {\ncr balance} between mutation and selection: alleles enter a population by mutation and, since most are deleterious, they are removed by directional selection
\myvs
\item Equality in the rate of addition due to mutation and removal due to selection is achieved
\myvs
\item First, approximate the change in $p$ in a single generation due to mutation:
\vspace{0.5in}
\item Then, derive an approximation due to selection:
\vspace{0.5in}
\item The {\ncb equilibrium frequency} is then:
\vspace{0.5in}
\end{itemize}

## \small Testing for selection: Tajima's D

\begin{itemize}
\item {\ncr Recall} our estimators of $\theta$: Watterson's estimator and Tajima's estimator
\myvs
\item Both are {\ncb unbiased} estimators of $\theta$ and so when they differ they suggest selection -- why?
\end{itemize}
\vspace{2.0in}

## \small Tajima's D statistic

\begin{itemize}
\item When the population is evolving neutrally, both the nucleotide polymorphism and the nucleotide diversity are {\ncr unbiased estimators} of $\theta$
\myvs
\item When their observed values are different, {\ncb deviation from neutrality} is suggested
\myvs
\item Tajima (1989) proposed a {\ncr test of neutrality} based on what is known as {\ncb Tajima's D statistic:}
$$ D = \frac{\hat{\theta}_T - \hat{\theta}_W}{\sqrt{c_1S + c_2 S (S-\frac{1}{k})}}$$
where $c_1 = \frac{b_1}{a_1} - \frac{1}{a_1^2}$ and $c_2 = \biggl(\frac{1}{a_1^2+a_2}\biggr) \biggl(b_2-\frac{n+2}{a_1n}+\frac{a_2}{a_1^2}\biggr)$
\myvs
\item When the null hypothesis of equilibrium and selective neutrality is true, $D$ has mean 0 and variance 1. P-values can be obtained using simulation.
\myvs
\item $D<0$ is often an indication of a {\ncr selective sweep}
\myvs
\item $D>0$ is often an indication of {\ncb balancing selection}
\end{itemize}
\vspace{0.2in}

## \small Tajima's D statistic

\begin{itemize}
\item {\ncr Example:} Rhodopsin 3 (Rh3) gene in {\itshape Drosophila simulans} (Ayala et al. 1993)
\end{itemize}

\vspace{0.in}

\scriptsize
\begin{tabular}{c | c c c c c c c c c c c c c c c c}
\hline
 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16\\
\hline
$a$ & T & C & T & A & C & C & T & C & C & T & C & G & G & T & T & A\\
$b$ & T & C & C & T & A & C & C & T & C & C & T & G & G & T & T & T\\
$c$ & C & T & C & C & C & C & C & T & C & T & T & T & G & C & T & A \\
$d$ & C & T & C & C & C & C & C & T & T & C & T & G & A & C & T & T \\
$e$ & C & T & C & C & C & T & C & T & T & T & T & G & G & C & C & A\\
\hline
& 6 & 6 & 4 & 7 & 4 & 4 & 4 & 4 & 6 & 6 & 4 & 4 & 4 & 6 & 4 & 6\\
\end{tabular}
\vspace{1.5in}
