---
title: "Species distribution modeling for ked in Ohio"
output: html_document
author: Week 14 Lab, EEOB 5330, Spring 2026
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document gives a basic sketch for running an ecological niche model (ENM). It considers ked (<i>Lipoptena cervi</i>) in Ohio. The steps required are below.

<br><br>


#### 0. Set up our R environment

We begin by installing necessary packages and creating a workspace.

```{r, echo=T} 
#install.packages(c("terra", "maxnet","pROC"))

library(terra)
library(maxnet)
library(pROC)
library(dismo)

dir.create("data", showWarnings = FALSE)
dir.create("data/bioclim", showWarnings = FALSE)
dir.create("data/studyarea", showWarnings = FALSE)
dir.create("output", showWarnings = FALSE)
```
<br><br>

#### 1. Download the climate data

We begin by downloading the climate data that will be used to predict the range of ked. The data are downloaded from worldclim.org, and include the following variables:

```{r table2, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
| Variable Name | Description   |
|---------------|:-------------:|
| BIO1          | Annual Mean Temperature (degrees Celsius times 10) |
| BIO2          | Mean Diurnal Range (Mean of monthly (max temp - min temp)) |
| BIO3          | Isothermality (BIO2/BIO7) (×100) |
| BIO4          | Temperature Seasonality (standard deviation ×100) |
| BIO5          | Max Temperature of Warmest Month |
| BIO6          | Min Temperature of Coldest Month |
| BIO7          | Temperature Annual Range (BIO5-BIO6) |
| BIO8          | Mean Temperature of Wettest Quarter |
| BIO9          | Mean Temperature of Driest Quarter |
| BIO10         | Mean Temperature of Warmest Quarter |
| BIO11         | Mean Temperature of Coldest Quarter |
| BIO12         | Annual Precipitation |
| BIO13         | Precipitation of Wettest Month |
| BIO14         | Precipitation of Driest Month |
| BIO15         | Precipitation Seasonality (Coefficient of Variation) |
| BIO16         |  Precipitation of Wettest Quarter |
| BIO17         | Precipitation of Driest Quarter |
| BIO18         | Precipitation of Warmest Quarter |
| BIO19         | Precipitation of Coldest Quarter |
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```
<br><br>

The climate data can be plotted, as below. Later, we will subset the data for Ohio to match the occurrence records for ked.

```{r, echo=T}
clim_files <- list.files("data/bioclim/", pattern = ".tif$", full.names = TRUE)
clim <- rast(clim_files)

plot(clim)
```
<br><br>

#### 2. Retrieve and display occurrence data

We next download occurrence data for ked (<i>Lipoptena cervi</i>) in Ohio. We clean the data by removing records for which latitude and/or longitude are missing or for which there are duplicated observations. We also remove observations collected before 1980. The output below gives characteristics of the data following cleaning.

If you were running this from scratch, you would pull the data from GBIF. Since this can take awhile, I've provide the results of this in the file <font color=blue>ked_raw.</font> Be sure that you have placed this in the proper directory. 

```{r, echo=T}
if(file.exists("data/ked_raw")){
  load("data/ked_raw")
}else{
  ked_raw=gbif(genus='Lipoptena',species='cervi') # only do this once -- this downloads data from gbif; read it in after this
  save(ked_raw,file = "data/ked_raw")
  write.csv("data/ked_raw.csv")
}

ked_clean <- subset(ked_raw, !is.na(lat) & !is.na(lon))
ked_unique <- ked_clean[!duplicated(ked_clean[, c("lat", "lon")]), ]
ked_unique <- subset(ked_unique, continent == "NORTH_AMERICA")

# check subsetting
table(ked_unique$year)
table(ked_unique$adm1)

ked_vect <- vect(ked_unique, geom = c("lon", "lat"), crs = "EPSG:4326")

# extract climate data from the locations where we have ked data
env_vals <- extract(clim, ked_vect)
ked_vect <- ked_vect[!is.na(env_vals[,2])]
```
<br><br>

We next make the occurrence data spatial for the records restricted to Ohio. The map below shows the occurrence data overlaid on the state of Ohio.

```{r, echo=T}
Ohio <- vect("data/GISpolygon/states.shp")
#Ohio <- states[states$coun == "Ohio"]

# check that Ohio is set up correctly; plot occurrence data
plot(Ohio)
points(ked_unique$lon,ked_unique$lat,pch=20)
```

<br><br>

The code below makes sure that there is only one observation per cell. It also extends the area considered a bit beyond Ohio's borders to help with making predictions at the edges. We then show plots of some of the bioclimatic variables limited to Ohio.

```{r, echo=T}

# thin points (one per raster cell)
cells <- cellFromXY(clim[[1]], geom(ked_vect)[,1:2])
ked_vect <- ked_vect[!duplicated(cells)]

# buffer
ked_buff <- buffer(ked_vect, width = 400000)

# crop + mask
ohio_ext <- ext(Ohio)

# expand by 1 degree in all directions
ohio_ext_expanded <- ext(
  xmin(ohio_ext) - 1,
  xmax(ohio_ext) + 1,
  ymin(ohio_ext) - 1,
  ymax(ohio_ext) + 1
)

studyArea <- crop(clim, ohio_ext_expanded)
studyArea <- mask(studyArea, ked_buff)


# now plot some of the bioclimatic variables
plot(studyArea[[1]],main="Annual Mean Temperature") #  note that units are degrees Celsius times 10
plot(Ohio,add=T)

plot(studyArea[[14]],main="Precipitation of driest month") 
plot(Ohio,add=T)

plot(studyArea[[9]],main="Mean temperature of driest quarter") 
plot(Ohio,add=T)

plot(studyArea[[11]],main="Mean Temperature of Coldest Quarter") 
plot(Ohio,add=T)

plot(studyArea[[4]],main="Temperature Seasonality (standard deviation ×100)") 
plot(Ohio,add=T)

plot(studyArea[[6]],main="Min Temperature of Coldest Month") 
plot(Ohio,add=T)

plot(studyArea[[2]],main="Mean Diurnal Range (Mean of monthly (max temp - min temp))") 
plot(Ohio,add=T)

```
<br><br>



#### 3. Prepare to run MaxENT

We will include some background points in our analysis. The code below specifies that we randomly pick 500 points across our study area.

```{r, echo=T}

set.seed(1)

bg <- spatSample(
  studyArea,
  size = 500,
  method = "random",
  na.rm = TRUE,
  as.points = TRUE
)

```

<br><br>

Next we divide our data into training and test sets, and combine our background and training data.

```{r, echo=T}
set.seed(1)

idx <- sample(1:nrow(ked_vect), nrow(ked_vect) * 0.5)

ked_train <- ked_vect[idx]
ked_test  <- ked_vect[-idx]

p <- extract(clim, ked_train)
a <- extract(clim, bg)

# remove ID column
p <- p[,-1]
a <- a[,-1]

# combine
X <- rbind(p, a)

# response vector
y <- c(rep(1, nrow(p)), rep(0, nrow(a)))

```

<br><br>

#### 4. Fit MaxENT and make predictions

Finally, we're ready to run our ENM model! Note that several modeling choices have been made.  Once the model is fit, we plot our predictions.

```{r, echo=T}

# run MaxENT
mod <- maxnet(
  p = y,
  data = X,
  f = maxnet.formula(y, X, classes = "lqph")  # linear, quadratic, product, hinge
)

# make predictions
pred <- predict(studyArea, mod, type = "cloglog", na.rm=TRUE)

plot(pred, main = "Predicted Ked Suitability")
plot(Ohio, add = TRUE)
points(geom(ked_vect), pch = 20)


plot(pred, main = "Predicted Ked Suitability")
plot(Ohio, add = TRUE)
points(geom(ked_vect), pch = 20)
```

<br><br>

#### 5. Evaluate predictions

We can evaluate the model using our test data.  This is an important step to calibrate how well our model works for data that it was not trained on. We want the Area Under the Curve (AUC) to be as large as possible.

```{r, echo=T}

# extract test data
p_test <- extract(clim, ked_test)[,-1]
a_test <- extract(clim, bg)[,-1]

test_data <- rbind(p_test, a_test)
test_labels <- c(rep(1, nrow(p_test)), rep(0, nrow(a_test)))

# predictions
pred_vals <- predict(mod, test_data, type = "cloglog")

# simple AUC
pred_vals <- as.vector(pred_vals)
auc(test_labels, pred_vals)

```

<br><br>

#### 6. Assess variable importance

We can use a permutation approach to assess variable importance. We shuffle the values of one variable and see how much predictions worsen (by looking at AUC). The larger the drop, the more important the variable. Values close to 0 indicate that the variable is not very important.

```{r, echo=T}

perm_importance <- function(model, X, y) {
  baseline <- predict(model, X, type = "cloglog")
  
  baseline_auc <- pROC::auc(y, as.vector(baseline))
  
  vars <- colnames(X)
  importance <- numeric(length(vars))
  
  for (i in seq_along(vars)) {
    X_perm <- X
    X_perm[, i] <- sample(X_perm[, i])  # shuffle
    
    pred_perm <- predict(model, X_perm, type = "cloglog")
    auc_perm <- pROC::auc(y, as.vector(pred_perm))
    
    importance[i] <- baseline_auc - auc_perm
  }
  
  names(importance) <- vars
  return(sort(importance, decreasing = TRUE))
}

library(pROC)

imp <- perm_importance(mod, X, y)
print(imp)

```

<br><br>


The variable importance values can also be plotted, which is a bit easier to look at!

```{r echo=T}

barplot(imp, las = 2, main = "Variable Importance (Permutation)",col="maroon")

```